---
title: "practical_exercise_1, Methods 3, 2021, autumn semester"
author: '[FILL IN YOUR NAME]'
date: "[FILL IN THE DATE]"
output: html_document
---

<style type="text/css">
  body{
  font-size: 14pt;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pacman::p_load(tidyverse)
```

# 3) Brushing up on the General Linear Model

We'll do a light start and get you back in the game of thinking about formulae and how to build your linear models  
Finally, we'll have a few exercises, finishing off today's practical exercises 

## A list of formulae
```{r, eval=FALSE}
formula <- y ~ x ## y as a function of x
y ~ 1 ## model the intercept for "y"
y ~ x ## model the main effect of x and the intercept for y
y ~ x + 1 ## the same as above (+ 1 is implicit)
y ~ x + 0 ## model the main effect of x and no intercept
y ~ x - 1 ## the same as above
y ~ 0 ## doesn't model anything (for completeness)
y ~ x + z ## model the main effects x and z (and an intercept)
y ~ x:z ## model interaction of x and z
y ~ x * z ## model the main effects x and z and their interaction
y ~ x + z + x:z ## the same as above
```

## Dataset mtcars
Let's look at the "mtcars" data:  

_[, 1]   mpg   Miles/(US) gallon  
[, 2]	 cyl	 Number of cylinders  
[, 3]	 disp	 Displacement (cu.in.)  
[, 4]	 hp	 Gross horsepower  
[, 5]	 drat	 Rear axle ratio  
[, 6]	 wt	 Weight (lb/1000)  
[, 7]	 qsec	 1/4 mile time  
[, 8]	 vs	 V/S  
[, 9]	 am	 Transmission (0 = automatic, 1 = manual)  
[,10]	 gear	 Number of forward gears  
[,11]	 carb	 Number of carburetors_  


## Miles per gallon and weight

We can do a scatter plot, and it looks like there is some relation between fuel usage and the weight of cars.
Let's investigate this further

```{r,fig.height=5, fig.width=6}
par(font.lab=2, font.axis=2, cex=1.2)
plot(mpg ~ wt, data=mtcars, xlab='Weight (lb/1000)', ylab='Miles/(US) gallon',
     main='Scatter plot', ylim=c(0, 40))
```

# Exercises and objectives
The objectives of today's exercises are:  
1) To remind you of the (general) linear model, and how we can use it to make models in R  
2) To make some informal model comparisons  
3) To estimate models based on binomially distributed data  

If you would like to read more about a given function, just prepend the function with a question mark, e.g.  
``` {r, eval=FALSE}
?lm
```

REMEMBER: In your report, make sure to include code that can reproduce the answers requested in the exercises below   

## Exercise 1
The general linear model: $Y = X \beta + \epsilon$:  
Do a linear regression, expressing fuel usage as a function of weight using the function __lm__  
```{r, eval=FALSE}
data(mtcars)
model <- lm(mpg ~ wt, data = mtcars)

summary(model)
model.matrix(model)
```
1. extract $\hat{\beta}$, $Y$, $\hat{Y}$, $X$ and $\epsilon$ from __model__ (hint: have a look at the function __model.matrix__)  
    i. create a plot that illustrates $Y$ and $\hat{Y}$ (if you are feeling ambitious, also include $\epsilon$ (hint: you can use the function __arrows__))
2. estimate $\beta$ for a quadratic model ($y = {\beta}_{2} x^2 + {\beta}_{1} x + {\beta}_{0}$) using ordinary least squares _without_ using __lm__; $\hat{\beta} = {({X}^{T} X)}^{-1} {X}^{T} Y$ (hint: add a third column to $X$ from step 1)
3. compare your acquired $\hat{\beta}$ with the output of the corresponding quadratic model created using __lm__ (hint: use the function __I__, see details under help and the sub-section formula operators here: https://www.datacamp.com/community/tutorials/r-formula-tutorial)  
    i. create a plot that illustrates $Y$ and $\hat{Y}$ (if you are feeling ambitious, also include $\epsilon$ (hint: you can use the function __arrows__)) 

### Answers
```{r}
###### 1.1 - extracting values and plotting ###### 
b_hat = model$coefficients
y = mtcars$mpg
X = cbind(1,mtcars$wt)
y_hat = X%*%b_hat
error = sigma(model)

df <- tibble(x = mtcars$wt, y = mtcars$mpg, y_hat = y_hat)

ggplot(data = df)+
  geom_smooth(aes(x, y, color = "data")) + # A line fittet to actual data-points
  geom_smooth(aes(x, y_hat, color = "model")) # The linear model based on points


######  1.2 - extracking b_hat using OLS ###### 
df$x2 <- df$x^2 # adding x^2 column to data-frame
X2 <- cbind(1, x = df$x, x2 = df$x2) # creating design matrix for quadratic model

b_hat2<- solve(t(X2)%*%X2)%*%t(X2)%*%y # extracting b_hat 
b_hat2

q_model <- lm(y ~ x + x2, data = df) # making a quadratic model using lm
summary(q_model)  

tibble(b_hat2, q_model$coefficients) # comparing coefficients extracted by hand with from lm-model


######  1.3 Plotting q_model and model Y and Y-hat  ###### 

# Finding Y_hat 
y_hat2 <- X2%*%b_hat2

df$y_hat2 <- y_hat2 # Adding the values to dataframe

# Plotting Y against Y_hat for quadratic model
ggplot(data = df)+
  geom_smooth(aes(x, y, color = "data")) +
  geom_smooth(aes(x, y_hat2, color = "model")) 

```

## Exercise 2
Compare the plotted quadratic fit to the linear fit  

1. which seems better?  
2. calculate the sum of squared errors, (show the calculation based on $\epsilon$). Which fit has the lower sum?  
3. now make a cubic fit ($y = {\beta}_{3} x^3 + {\beta}_{2} x^2 + {\beta}_{1} x + {\beta}_{0}$) and compare it to the quadratic fit  
    i. create a plot that illustrates $Y$ and $\hat{Y}$ for both the cubic and the quadratic fits (plot them in the same plot)  
    ii. compare the sum of squared errors  
    iii. what's the estimated value of the "cubic" (${\beta}_3$) parameter? Comment on this!  
4. bonus question: which summary statistic is the fitted value (_Intercept_ or ${\beta}_0$ in $y = {\beta}_0$) below identical to?
```{r, echo=FALSE}
lm(mpg ~ 1, data=mtcars)
```

# Exercise 2 answers 



```{r}
######  2.1 Comparing models ###### 
# Linear model 
ggplot(data = df)+
  geom_smooth(aes(x, y, color = "data")) + 
  geom_smooth(aes(x, y_hat, color = "model")) 

# Quadratic model
ggplot(data = df)+
  geom_smooth(aes(x, y, color = "data")) +
  geom_smooth(aes(x, y_hat2, color = "model")) 

summary(model) # R^2 = 0.7446
summary(q_model) # R^2 = 0.8066

# The quadratic fit seems to fit more of the data, both looking at the plots and comparing R^2 values. The quadratic model explains 81% of variance and the linear model explains 74%.


######  2.2 calculate the sum of squared errors: SSE = Σ(ŷi – yi)2 ###### 

sse <- sum((y_hat - y)^2) # SSE for linear model
sse_q <- sum((y_hat2 - y)^2) # SSE for quadratic model
(sse-sse_q)/sse # difference in percentage
# We see that the SSE value dropped with around 27% by making our model quadratic


######  2.3 Cubic fits ###### 

# We do that by cubing the predictor
df$x3 <- df$x^3

# Building the model
c_model = lm(y ~ x3 + x2 + x, data = df)
summary(c_model)

# Adding y_hat to dataframe
X3 <- model.matrix(c_model) # creating design matrix for cubic model
y_hat3 <- X3%*%coef(c_model) # finding y_hat for cubic model
df$y3 <- y_hat3

# 2.3 i: Plotting cubic and quadratic models
ggplot(data = df) +
  geom_smooth(aes(x = x, y = y, colour = "Data"))+
  geom_smooth(aes(x = x, y = y_hat2, colour = "Quad")) +
  geom_smooth(aes(x = x, y = y_hat3, colour = "Cubic"))

# 2.3 ii: Comparing SEE
sse_cubic <- sum((y_hat3 - y)^2)
sse_cubic
# The SSE for the cubic model is SLIGHTY better than for the quadratic model 

# 2.3 iii: Commenting on the estimated B_hat value 
coef(c_model)[2] # x3 = 0.046
# The fourth parameter in the cubic model (0.046) is very small, meaning that the cubic element adds very little information 
```

## Exercise 3
Doing a logistic regression - estimating the probability that a car has automatic transmission (0) or manual transmission (1) based on its weight
```{r}
logistic.model <- glm(am ~ wt, data = mtcars, family="binomial")
summary(logistic.model)

model.matrix(logistic.model)

# Finding estimated values 
fitted_values <- fitted(logistic.model)
linear.predicters = inv.logit(logistic.model$coefficients[1] + logistic.model$coefficients[2]*mtcars$wt)
# The fitted values and linera predictors are identical 


# Plotting the logistic function
df_log <- tibble(x = mtcars$wt, y = fitted_values)
ggplot(df_log, aes(x, y)) +
  geom_line(xlim = c(0, 7))

# 3.2 i. interpreting the intercept

boot::inv.logit(logistic.model$coefficients[1]) # First we take the inverse of the intercept value
# This is .99
# This means that if a car weighs 0 then the probability of it being manual gear is 99%

# 3.2 ii. probability of the Pontiac Firebird has automatic transmission?
# First we find the weight of the car
mtcars %>% 
  filter(row.names(mtcars) == "Pontiac Firebird") %>%
  select(wt) # = 3.845

# Then we multiply the weight with slope and add the intercept
boot::inv.logit(logistic.model$coefficients[1] + 3.845*logistic.model$coefficients[2])
# There's a 3.13 % probability that the Pontiac Firebird 

# 3.3 quadratic fit
mtcars$wt2 <- mtcars$wt^2 # adding the quadratic term to dataframe
logistic.model_q <- glm(am ~ wt2 + wt, data = mtcars, family="binomial")
summary(logistic.model_q)

plot(mtcars$wt, logistic.model$coefficients[1] + logistic.model$coefficients[2]*mtcars$wt)
plot(mtcars$wt, logistic.model_q$coefficients[1] + logistic.model_q$coefficients[2]*mtcars$wt2)



# Comparing 
AIC(logistic.model, logistic.model_q)
  
```

Probabilities live on the range $(0, 1)$ - using the so-called logit function as a "link-function" we can map these onto the range $(-\infty, \infty)$, i.e. the real numbers.  
  
What we model in this case is: $Pr(y = 1) = logit^{-1}(X \beta)$, i.e. the probability of a car having manual transmission, given its weight. $X \beta$ is called the linear predictor; compare with $Y = X \beta + \epsilon$ 
It is helpful to define the logit function and its inverse function for the following:  

```{r}
logit <-     function(x) log(x / (1 - x))
inv.logit <- function(x) exp(x) / (1 + exp(x))
```

1. plot the fitted values for __logistic.model__:  
    i. what is the relation between the __linear.predictors__ and the __fitted_values__ of the __logistic.model__ object?
2. plot the logistic function, you've estimated based on your $\hat{\beta}$, (not just the fitted values). Use an _xlim_ of (0, 7)
    i. what's the interpretation of the estimated $\hat{\beta}_0$ (the _Intercept_)
    ii. calculate the estimated probability that the Pontiac Firebird has automatic transmission, given its weight
    iii. bonus question - plot the logistic function and highlight all the cars where we guessed wrongly, if we used the following "quantizer" function:
    
\begin{equation}
  transmission_{guess}=
  \begin{cases}
    1 (manual), & \text{if}\ PR(y = 1) ≥ 0.5 \\
    0 (automatic), & \text{otherwise}
  \end{cases}
\end{equation}    
    
3. plot quadratic fit alongside linear fit  
    i. judging visually, does adding a quadratic term make a difference?
    ii. check the details in the help of the AIC function - which of the models provide the better fit according to the AIC values and the residual deviance respectively?
    iii. in your own words, why might it be good to penalise a model like the quadratic model, we just fitted.
    
# Next time
We are going to looking at extending our models with so called random effects. We need to install the package "lme4" for this. Run the code below or install it from your package manager (Linux)  
```{r, eval=FALSE}
install.packages("lme4")
```
We can fit a model like this:

```{r}
library(lme4)
mixed.model <- lmer(mpg ~ wt + (1 | cyl), data=mtcars)
```

They result in plots like these:
```{r}
par(font.lab=2, font.axis=2, cex=1.2)
plot(mtcars$wt, fitted.values(mixed.model),
     main='Linear regression with group intercepts (n cylinders)',
    xlab='Weight (lb/1000)', ylab='Miles/(US) gallon',
    pch=3)
```

and this
```{r}
mixed.model <- lmer(mpg ~ wt + (wt | cyl), data=mtcars)
plot(mtcars$wt, fitted.values(mixed.model),
     main='Linear regression with group intercepts and group slopes (n cylinders)',
    xlab='Weight (lb/1000)', ylab='Miles/(US) gallon',
    pch=3)
``` 

but also new warnings like:  

Warning:
In checkConv(attr(opt, "derivs"), opt\$par, ctrl = control$checkConv,  :
  Model failed to converge with max|grad| = 0.0121962 (tol = 0.002, component 1)
